|

### 2024-10-21 manuscript/paper.Rmd
manuscript improvements, text, revisions, edits.

---
title: |
  |  \Large Increased sexual dimorphism evolves in a fossil stickleback following ecological release from fish piscivores
author: | 
  | \large Matthew Stuart$^{1,2}$, Allison Ozark$^{3}$, Raheyma Siddiqui$^3$, Akhil Ghosh$^{1,2}$
  | \large Samantha Swank$^3$, Michael A. Bell$^4$, Gregory J. Matthews$^{1,2}$, and Yoel E. Stuart$^{3,+}$ 
  | \vspace{-1.1mm}
  | \large $^1$ Department of Mathematics and Statistics, Loyola University Chicago, Chicago, IL, USA \vspace{-1.1mm}
  | \large $^2$ Center for Data Science and Consulting, Loyola University Chicago, Chicago, IL, USA \vspace{-1.1mm}
  | \large $^3$ Department of Biology, Loyola University Chicago, Chicago, IL, USA \vspace{-1.1mm}
  | \large $^4$ University of California Museum of Paleontology, Berkeley, CA, USA \vspace{-1.1mm}
  | \large $^+$ Corresponding: ystuart@luc.edu \vspace{-1.1mm}
abstract: |
  | Everyone loves the stickle \vspace{2mm}
  | *Keywords*: Stickle
bibliography: references.bib
fontsize: 12pt
link-citations: true
linkcolor: cyan
urlcolor: cyan
output:
  pdf_document:
    df_print: kable
    number_sections: true
    keep_tex: true
header-includes:
 \usepackage{setspace}
 \setstretch{1.15}
 \usepackage{float}
 \floatplacement{figure}{t}
---

@@

| Trait Name  | Trait Code | Trait Description           |
|---------|-----|------------------------------------------|
|Standard Length|	`stl` |	Distance from anterior tip of premaxilla to posterior end of last vertebra (hypural plate)|
|Dorsal Spine | `mds` |	Number of dorsal spines from 0 to 3|
|Dorsal Fin Ray|`mdf`|	Number of bones in the dorsal fin posterior to the third dorsal spine (i.e., soft dorsal fin rays)
|Anal Fin Ray |`maf`|	Number of bones in the anal fin posterior to the anal spine (i.e., soft dorsal fin rays)|
|Abdominal Vertebra|mav|	Number of vertebrae anterior to the first vertebra contacting an anal fin pterygiophore (Aguirre et al. 2014)|
|Caudal Vertebra|`mcv`	| Number of vertebrae posterior to and including the first vertebra contacting an anal fin pterygiophore (Aguirre et al. 2014)|
|Pterygiophore number|`mpt`|	Number of pterygiophores anterior to but excluding the pterygiophore under the third dorsal spine, which is immediately anterior to and contiguous with the dorsal fin|
|Pelvic Spine length|`lps.sc`	|Length from the base of one pelvic spine above its articulation with the pelvic girdle to its distal tip|
|Ectocoracoid|`ect.sc`|	Length between the anterior and posterior tips of the shoulder girdle base (i.e., ectocoracoid)|
|Pelvic Girdle|`tpg.sc`|	Length between the anterior to posterior tips along midline. If vestigial, the sum of longest anterior-posterior axis for the vestiges|
|Cleithrum length|`cle.sc`|	Length from free dorsal tip to ventral tip of the cleithrum on the anterior margin of the shoulder girdle (i.e., cleithrum)|
|Premaxilla|`pmx.sc`|	Length from the anterior tip of the premaxilla to the distal tip of the ascending process of the premaxilla |
|Dorsal Spine|`Ds#.sc #=1/2/3`|	Length from the base of a dorsal spine above the pterygiophore to its distal tip along the anterior edge|
|Pterygiophore|`lpt.sc`|	Distance between the anterior to posterior tips of the pterygiophore immediately preceding the 3rd dorsal spine (when present)|

@@

## Extant Specimen data
To span the gamut of stickleback diversity for our predictive imputation model, we sampled modern stickleback from lakes containing generalist stickleback populations (@Hendryetal2009; @Bolnick2011) and from lakes containing benthic-limnetic species pairs (@Baumgartneretal1988; @SchluterandMcPhail1992). The generalist specimens used here were collected by YES in 2013 and were previously described in @Stuartetal2017 (Table S1). These samples were fixed in formalin, then stained for bone with Alizarin Red in 2013. Benthic and limnetic specimens were kindly loaned by D. Schluter and S. Blain at the University of British Columbia. The Schluter lab collected benthic and limnetic individuals from Enos Lake in 1988 and from Emily Lake, Little Quarry Lake, Paxton Lake, and Priest Lake in 2018 (Table S1). The Enos specimens had been fixed whole in formalin and stored in 40% isopropanol. The specimens from the other lakes were initially preserved whole in 95% ethanol in the field before being gradually transferred to water then formalin in the lab and ultimately stored in 40% isopropanol. In 2019, we stained these specimens for bone using Alizarin Red.
We next replicated fossil data collection (Table 1) on these extant specimens. Standard length as well as pelvic-spine length on each side were measured with calipers. We used a dissection microscope to count dorsal spines, pelvic spines, dorsal-fin rays, and anal-fin rays. Right and left-side pelvic girdle lengths and ectocoracoid lengths were measured from ventral photographs taken using a Canon EOS Rebel T7 with a Tamron 16-300 mm MACRO lens mounted on a leveled Kaiser RS1 copy stand. Specimens were held in place for ventral photographs using a small tabletop vise with an attached scale bar. Lateral X-rays were used to measure dorsal spine length, number of pterygiophores anterior to the pterygiophore holding the third spine, length of the pterygiophore just anterior to the third spine, cleithrum length, and pre-maxilla ascending branch length. We also counted vertebrae from the X-rays: abdominal vertebrae were counted anterior to the first vertebra with a haemal spine contacting an anal fin pterygiophore. Caudal vertebrae were posterior, including the first vertebra with the haemal spine contacting the anal fin pterygiophore (following @Aguirreetal2014). X-rays were taken with an AXR Hot Shot X-ray Machine (Associated X-ray Corporation) at the Field Museum of Natural History. Specimens were exposed at 35kV and 4mA. Small fish were exposed for 7s, medium fish for 8s, and large fish for 10s. We developed the film and scanned individual images of each fish using the B&W Negatives setting on an Epson Perfection 4990 Photo flatbed at 2400 dpi. Measurements from photographs and X-rays were taken with FIJI (@Schindelinetal2012) and its plugin ObjectJ (https://sils.fnwi.uva.nl/bcb/objectj/). We dissected individuals from the generalist populations (Table S1) to determine sex from the gonads. Individuals from the species-pair lakes (TableS1) were previously sexed by the Schluter lab, using either dissection or a genotyping protocol \textcolor{red}{(Whom, personal communication)}.

@@

@@

We discuss the procedure for estimating this value below. \textcolor{red}{DID THIS GET DISCUSSED?-YS}

Priors: To aid in the sampling procedure for the discrete phenotypes, we first had to transform the parameters of the generalized Poisson distribution onto the real number line in order to ensure our algorithm can properly sample from the posterior distribution of interest using STAN software. This was done by declaring $\phi_k = \log\left(\frac{\alpha_k - \max_{i,j}(-\lambda_{k,ij}/y_{k,ij})}{1 - \alpha_k}\right)$ where $\lambda_{k,ij} = \left\{\begin{array}{cc} \exp(\mu_{k,ft_i} + \gamma_k Y_{K,ij}) & W_{ij} = \text{Female} \\ \exp(\mu_{k,mt_i} + \gamma_k Y_{K,ij}) & W_{ij} = \text{Male} \end{array}\right.$.

@@

We follow the method outlined in @ZhouandReiter to obtain $R$ draws from the posterior distribution of the outlined model parameters given our observed data and imputed values. More specifically, for the $j^{th}$ imputed data set where $j = 1,\cdots,M$, we obtain $\frac{R}{M}$ draws from the posterior distribution using an MCMC algorithm with a sufficient burnin period, and then we combine these $M$ MCMC draws to say that we have a total of $R$ draws from the overall posterior distribution of interest, which can then be used to assess the presence of sexual dimorphism for a given trait.

After running these models, we have posterior distributions of all the model parameters of interest. From these, we can obtain samples from the posterior distribution of mean trait values for each trait for males and females, using equation 6. We use these values for the following analyses.

## Quantifying sexual dimorphism, and testing for increased sexual dimorphism through time.
We identify sexual dimorphism first by calculating the difference in mean trait size for males and females at each time for phenotype $k$ ($\mu_{k,m,t_i} - \mu_{k,m,t_i}$) for $i = 1,\cdots,T$ where, for the continuous phenotypes, the mean trait sizes for each sex $\mu_{k,m,t_i}$ and $\mu_{k,f,t_i}$ are calculated using (\ref{eq:mu}).  For the discrete phenotypes, we note that in (\ref{eq:disc_corr}) and (\ref{eq:disc_ind}) the mean trait size is $\lambda_{k,m,t_i} = \exp(\mu_{k,m,t_i})$ and $\lambda_{k,f,t_i} = \exp(\mu_{k,f,t_i})$ for males and females, respectively, which is the value we use in our analysis. For simplicity, we will use $\mu_{k,m,t_i}$ and $\mu_{k,f,t_i}$ as the notation for mean size for both discrete and continuous traits for the rest of this discussion. From our MCMC algorithm, we obtain $M$ samples from the posterior distribution of $\mu_{k,m,t_i} - \mu_{k,f,t_i}$ to use in our analysis.

Recall that we predicted that sexual dimorphism should increase through time in Lineage II, following ecological release from predators. However, this change in dimorphism make be increasing at a non-constant rate, as pointed out by the inclusion of the OU process. We assess for dimorphism by calculating a Kendall's $\tau$ coefficient (@KendallsTau). Kendall's $\tau$ assesses the ordinal association between a set of bivariate observations, where $\tau > 0$ indicates concordance between the observations (i.e., as $x$ increases, $y$ also tends to increase.)

For our analysis, we calculate $\boldsymbol{\tau}_k = \{\tau_{k,r}: r = 1,\cdots,R\}$ for trait $k$ where $\tau_r$ is the estimate of the Kendall's $\tau$ correlation for $\mu_{k,m,t_i} - \mu_{k,f,t_i}$ versus time $t_i$ for the $r^{th}$ MCMC sample from the posterior distribution. The posterior probability of sexual dimorphism for trait $k$ can be estimated as the proportion of values in $\boldsymbol{\tau}_k$ that are greater than 0 ($\frac{1}{R} \sum_{r=1}^R I(\tau_{k,r} > 0)$).  We can also assess the strength of the dimorphism by calculating the proportion of posterior samples that are greater than a prespecified threshold \textcolor{red}{(0.5??, idk what this could be.) -MS}

# Results {#sec:results}
\textcolor{red}{WE NEED RESULTS SECTION DESCRIBING THE RESULTS OF AKHIL'S MULTIPLE IMPUTATION, BOTH THE TESTS OF HOW WELL THE MODEL WORKED CLASSIFYING MODEL SAMPLES, AND SUMMARY STATISTICS OF FOSSIL IMPUTATION RESULTS -YS}
The extant data used here consists of a total of `r sum(!is.na(stick$gender))` specimens all with known sex (Table S1).  Of these, there are `r sum(stick$gender == "Female",na.rm = TRUE)` and `r sum(stick$gender == "Male",na.rm = TRUE)` female and male specimens, respectively.

@@

Traits such as standard length (stl), anal fin ray (maf), dorsal fin ray (mdf), premaxilla (pmx), and dorsal spine 1 (ds1) demonstrate constant sexual dimorphism with all but the last having learger mean values in male sepcimen. Other traits such as abdominal vertebra (mav), caudal vertebra (mcv), and pelvic spine length (lps) don't appear to have any discernable sexual dimorphism in the means.  Dorsal spine 3 (ds3) exhibits sexual dimorphism that changes over time.  The mean number of dorsal spine 3 begins larger for females for the earliest observed time periods.  Over time, however, this switches and the male specimens exhibit larger mean values for the latest observed time periods.

\textcolor{red} {
Yoel's interpretation of the three paragraphs above. Male and female G. doryssus differ for some traits, indicating detectable sexual dimorphism. This is evidenced by traits with substantial differences in male and female posterior means (i.e., deviation from 0 in Figure \ref{fig:mu-diff-plot}), supported by posterior probabilities that deviate from 0.5 (Figure \ref{fig:prob0-plot}). For example, males are larger than females by 3-8mm on average through time (stl panel in Figure \ref{fig:mu-diff-plot}), with a posterior probability for that mean difference above 70 percent through time (Figure \ref{fig:prob0-plot}). In contrast, the number of abdominal vertebrae (mav) fluctuates about 0 (Figure \ref{fig:mu-diff-plot}), with posterior probabilities fluctuating about 50 percent through time (Figure \ref{fig:prob0-plot}). Table \ref{tab:prob0-table} in the Supplementary Material summarizes the posterior probabilities that male means differ from female means by trait and sample.
}

\textcolor{red}{can we move mu-diff-plot and prob0-plot here?}

## Changes over time
Table \ref{tab:auc-table} shows the probability that the posterior distribution of the differences in the mean is greater than the posterior distribution of the differences at the first time point.  Values near 0.5 indicate very little difference between the distributions whereas values closer to 0 and 1 indicate changes in the distributions with values of 1 meaning the difference in means has shifted towards males having a larger mean and values of 0 indicating the difference in means has shifted towards larger means in females.  These results are shown graphically in figure \ref{fig:AUC-plot}.

The amount of sexual dimporphism over time does not change substantially for a majority of the trait observed here (i.e. cle, ds1, ds2, ect, lps, lpt, mav, mcv, pmx, stl).  Traits that trend away from the mean difference at the first observed time point include dorsal spine 3 (ds3), anal fin ray (maf), and dorsal fin ray (mdf).

<!-- #ds3 -->
<!-- apply(mu_m,2,mean)[1] #2.291846 -->
<!-- apply(mu_f,2,mean)[1] #2.448115  -->
<!-- apply(mu_m,2,mean)[18] #1.264728  -->
<!--   apply(mu_f,2,mean)[18] #1.184534  -->
<!-- apply(mu_m,2,mean)[18] - apply(mu_m,2,mean)[1] #-1.027118  -->
<!-- apply(mu_f,2,mean)[18] - apply(mu_f,2,mean)[1] #-1.263581  -->
The posterior mean of the mean number of dorsal spine rays at the first observed time point is 2.29 for males and for females it is 2.45.  At the last observed time point, the average number for males and female both dropped to 1.265 and 1.185 for males and females, respectively.  This resulted in a mean number of dorsal spine 3 lost by males of 1.027 and for females it was 1.264.  This resulted in a change of the mean difference from time point 1 to time point 18 of 0.236.  Basically, males and females both lowered the mean number of dorsal spine 3, but females lowered their mean more than the males.

@@

We also observed a change in the differences between males and females for the number of bones int he anal fin ray (maf) and the number of bones in the dorsal fin ray.  The difference in the means of males versus females for anal fin ray bones changed by 0.152 over the course of the observed time periods.  This change was driven mostly by the change in females.  At the first observed time period the had on average 8.291 bones dropping to 8.170 over the 18 times periods observed. Whereas for males the change in means rose only from  8.518 to 8.549, about 0.0316 bones on average.

A similar change in the difference of the means over time was observed for the number of bones in the dorsal fin (mdf).  However, unlike maf, this change in sexual dimporhism was driven entirely by the males.  The mean change for females over the 18 observed time periods was only 0.0055 with the average number of dorsal fin bones for females starting at 9.287 and the final observed mean of 9.281 resulting in almost no change in the mean.  For males, the average number of dorsal fin bones at the first time point was about 9.568 dropping to 9.734 over the course of the observed time periods for a gain of 0.166 bones on average.

@@

\textcolor{red} {
Yoel's interpretation of the five paragraphs above. 
Figure \ref{fig:AUC-plot} shows the posterior probability that the mean male-female difference at time t is different from the male-female difference at time zero. In other words, Figure \ref{fig:AUC-plot} plots whether the sexual dimorphism changes in subsequent samples, always relative to the first sample. The amount of sexual dimporphism over time does not change substantially for a majority of the traits (i.e. cle, ds1, ds2, ect, lps, lpt, mav, mcv, pmx, stl; Figure \ref{fig:AUC-plot}). Traits that change in dimorphism are anal fin ray (maf), and dorsal fin ray (mdf), and dorsal spine 3 (ds3).}

>>>>

---
title: |
  |  \Large Increased sexual dimorphism does not evolve in a fossil stickleback following ecological release from fish piscivores
author: | 
  | \large Matthew Stuart$^{1,2}$, Allison Ozark$^{3}$, Raheyma Siddiqui$^3$, Akhil Ghosh$^{1,2}$
  | \large Samantha Swank$^3$, Michael A. Bell$^4$, Gregory J. Matthews$^{1,2}$, and Yoel E. Stuart$^{3,+}$ 
  | \vspace{-1.1mm}
  | \large $^1$ Department of Mathematics and Statistics, Loyola University Chicago, Chicago, IL, USA \vspace{-1.1mm}
  | \large $^2$ Center for Data Science and Consulting, Loyola University Chicago, Chicago, IL, USA \vspace{-1.1mm}
  | \large $^3$ Department of Biology, Loyola University Chicago, Chicago, IL, USA \vspace{-1.1mm}
  | \large $^4$ University of California Museum of Paleontology, Berkeley, CA, USA \vspace{-1.1mm}
  | \large $^+$ Corresponding: ystuart@luc.edu \vspace{-1.1mm}
abstract: |
  | Theory suggests that populations should expand their habitat- and resource-use niches when they are freed from interactions with competitors and predators--so-called ecological release. One way that this ecological release can manifest is through increased differences between males and females of the same species as they diverge into different niches. If divergent niche use drives corresponding divergent adaptation in in the traits used to exploit the divergent niches, then theory predicts an increase in sexual dimorphism for those traits, sometimes called character displacement between the sexes. Using a dataset of 16 traits collected from 18 temporal samples spread across 16,500 years of evolution of the fossil stickleback fish, *Gasterosteus doryssus*, we tested the prediction that release from predation inferred in this system would result in character displacement between the sexes. We used data from populations of extant stickleback (*Gasterosteus aculeatus*) to build an imputation model to classify individual sex in fossil specimens. Then, we quantified the extent of sexual dimorphism at each temporal sample, and tracked how it changed through time. NEED TO FINISH THIS   \vspace{2mm}
  | *Keywords*: character displacement, character release, evolutionary time series, *Gasterosteus doryssus*, missing data imputation
bibliography: references.bib
fontsize: 12pt
link-citations: true
linkcolor: cyan
urlcolor: cyan
output:
  pdf_document:
    df_print: kable
    number_sections: true
    keep_tex: true
header-includes:
 \usepackage{setspace}
 \setstretch{1.15}
 \usepackage{float}
 \floatplacement{figure}{t}
---

@@

| Trait Name  | Trait Code | Trait Description           |
|------------|-----|------------------------------------------|
|Standard Length|	`stl` |	Distance from anterior tip of premaxilla to posterior end of last vertebra (i.e., the llhypural plate)|
|Dorsal Spine Count| `mds` |	Number of dorsal spines|
|Dorsal Fin Ray Count|`mdf`|	Number of bones in the dorsal fin posterior to the third dorsal spine (i.e., soft dorsal fin rays)
|Anal Fin Ray Count|`maf`|	Number of bones in the anal fin posterior to the anal spine (i.e., soft anal fin rays)|
|Abdominal Vertebra Count|mav|	Number of vertebrae anterior to the first vertebra contacting an anal fin pterygiophore (Aguirre et al. 2014)|
|Caudal Vertebra Count|`mcv`	| Number of vertebrae posterior to and including the first vertebra contacting an anal fin pterygiophore (Aguirre et al. 2014)|
|Pterygiophore Count|`mpt`|	Number of pterygiophores anterior to but excluding the pterygiophore under the third dorsal spine|
|Pelvic Spine Length|`lps.sc`	|Length from the base of one pelvic spine at its pelvic girdle articulation to its distal tip|
|Ectocoracoid Length|`ect.sc`|	Length from the anterior to posterior tips of the shoulder girdle base|
|Pelvic Girdle Length|`tpg.sc`|	Length from the anterior to posterior tips along midline. If vestigial, the sum of longest axis along vestiges|
|Cleithrum Length|`cle.sc`|	Length from dorsal tip to ventral tip on the anterior margin of the shoulder girdle|
|Premaxilla Length|`pmx.sc`|	Length from the anterior tip of the premaxilla to the distal tip of its ascending process|
|Dorsal Spine Length|`Ds#.sc #=1/2/3`|	Length from the dorsal spine base at the pterygiophore to its distal tip|
|Pterygiophore Length|`lpt.sc`|	Distance from anterior to posterior edges of the pterygiophore immediately before ds3 (when present)|

@@

## Extant Specimen data
To span the gamut of stickleback diversity for our predictive imputation model, we sampled modern stickleback from lakes containing generalist stickleback populations (@Hendryetal2009; @Bolnick2011) and from lakes containing benthic-limnetic species pairs (@Baumgartneretal1988; @SchluterandMcPhail1992). The generalist specimens used here were collected by YES in 2013 and were previously described in @Stuartetal2017 (Table S1). These samples were fixed in formalin, then stained for bone with Alizarin Red in 2013. Benthic and limnetic specimens were kindly loaned by D. Schluter and S. Blain at the University of British Columbia. The Schluter lab collected benthic and limnetic individuals from Enos Lake in 1988 and from Emily Lake, Little Quarry Lake, Paxton Lake, and Priest Lake in 2018 (Table S1). The Enos specimens had been fixed whole in formalin and stored in 40% isopropanol. The specimens from the other lakes were initially preserved whole in 95% ethanol in the field before being gradually transferred to water then formalin in the lab and ultimately stored in 40% isopropanol. In 2019, we stained these specimens for bone using Alizarin Red.

We next replicated fossil data collection (Table 1) on these extant specimens. Standard length as well as pelvic-spine length on each side were measured with calipers. We used a dissection microscope to count dorsal spines, pelvic spines, dorsal-fin rays, and anal-fin rays. Right and left-side pelvic girdle lengths and ectocoracoid lengths were measured from ventral photographs taken using a Canon EOS Rebel T7 with a Tamron 16-300 mm MACRO lens mounted on a leveled Kaiser RS1 copy stand. Specimens were held in place for ventral photographs using a small tabletop vise with an attached scale bar. Lateral X-rays were used to measure dorsal spine length, number of pterygiophores anterior to the pterygiophore holding the third spine, length of the pterygiophore just anterior to the third spine, cleithrum length, and pre-maxilla ascending branch length. We also counted vertebrae from the X-rays: abdominal vertebrae were counted anterior to the first vertebra with a haemal spine contacting an anal fin pterygiophore. Caudal vertebrae were posterior, including the first vertebra with the haemal spine contacting the anal fin pterygiophore (following @Aguirreetal2014). X-rays were taken with an AXR Hot Shot X-ray Machine (Associated X-ray Corporation) at the Field Museum of Natural History. Specimens were exposed at 35kV and 4mA. Small fish were exposed for 7s, medium fish for 8s, and large fish for 10s. We developed the film and scanned individual images of each fish using the B&W Negatives setting on an Epson Perfection 4990 Photo flatbed at 2400 dpi. Measurements from photographs and X-rays were taken with FIJI (@Schindelinetal2012) and its plugin ObjectJ (https://sils.fnwi.uva.nl/bcb/objectj/). We dissected individuals from the generalist populations (Table S1) to determine sex from the gonads. Individuals from the species-pair lakes (TableS1) were previously sexed by the Schluter lab, using either dissection or a genotyping protocol \textcolor{red}{(Whom, personal communication)}.

@@

\textcolor{red}{MATT, somewhere we need to describe how you handled traits with lots of evolved zeros (ds1, ds2, ds3, lps, tpg) -YS}

@@

\textcolor{red}{Q FOR MATT. Is transforming the GP onto the real number line separate from setting the priors? If so, should the statement of transformation go in the section where the GP and discrete trait modeling is discussed. Then priors get their own section here? -YS}
To aid in the sampling procedure for the discrete phenotypes, we first had to transform the parameters of the generalized Poisson distribution onto the real number line in order to ensure our algorithm can properly sample from the posterior distribution of interest using STAN software. This was done by declaring $\phi_k = \log\left(\frac{\alpha_k - \max_{i,j}(-\lambda_{k,ij}/y_{k,ij})}{1 - \alpha_k}\right)$ where $\lambda_{k,ij} = \left\{\begin{array}{cc} \exp(\mu_{k,ft_i} + \gamma_k Y_{K,ij}) & W_{ij} = \text{Female} \\ \exp(\mu_{k,mt_i} + \gamma_k Y_{K,ij}) & W_{ij} = \text{Male} \end{array}\right.$.

@@

We follow the method outlined in @ZhouReiter2010 to obtain $R$ draws from the posterior distributions of the model parameters given our observed data and imputed values. More specifically, for the $j^{th}$ imputed data set where $j = 1,\cdots,M$, we obtain $\frac{R}{M}$ draws from the posterior distribution using an MCMC algorithm with a sufficient burnin period, and then we combine these $M$ MCMC draws to say that we have a total of $R$ draws from the overall posterior distribution of interest.

For the continuous phenotypes, the mean trait sizes for each sex $\mu_{k,m,t_i}$ and $\mu_{k,f,t_i}$ are calculated using (\ref{eq:mu}).For the discrete phenotypes, equations (\ref{eq:disc_corr}) and (\ref{eq:disc_ind}) calculate the mean trait size is $\lambda_{k,m,t_i} = \exp(\mu_{k,m,t_i})$ and $\lambda_{k,f,t_i} = \exp(\mu_{k,f,t_i})$ for males and females. For simplicity, we will use $\mu_{k,m,t_i}$ and $\mu_{k,f,t_i}$ as the notation for mean size for both discrete and continuous traits for the rest of this discussion. From our MCMC algorithm, we obtain $M$ samples from the posterior distribution of $\mu_{k,m,t_i} - \mu_{k,f,t_i}$. That is, with posterior distributions of the model parameters described above, we obtained $M$ samples from the posterior distribution of mean trait values by trait for males and females at each time step. We then used these values to quantify the presence or absence of sexual dimorphism, and describe how sexual dimorphism evolved through time.

## Quantifying sexual dimorphism, and testing for increased sexual dimorphism through time.
We quantified sexual dimorphism at each time step by calculating the difference in mean trait size for males and females at each time for phenotype $k$, $\mu_{k,m,t_i} - \mu_{k,f,t_i}$, for $i = 1,\cdots,T$. The presence of sexual dimorphism was indicated by... \textcolor{red}{Matt, can you help me with Bayesian the language here? I think this is the place to reference the two post probs plots, yes? Values far from fifty fifty indicate that the mass of the posterior distribution is away from a male female difference of zero. YS}

Then, to test our prediction that ecological release should result in an increase in sexual dimorphism through time, we calculated Kendall's $\tau$ coefficient for each trait (@KendallsTau). Kendall's $\tau$ assesses the ordinal association between a set of bivariate observations, where $\tau > 0$ indicates concordance between the observations (i.e., as $x$ increases, $y$ also tends to increase; here,$x$ is time and $y$ is sexual dimorphism.) Kendall's $\tau$ varies between -1 and 1, and can be thought of as behaving like a correlation. Values farther from zero indicate a stronger relationship.

$\boldsymbol{\tau}_k = \{\tau_{k,r}: r = 1,\cdots,R\}$ for trait $k$ where $\tau_r$ is the estimate of the Kendall's $\tau$ correlation for $\mu_{k,m,t_i} - \mu_{k,f,t_i}$ versus time $t_i$ for the $r^{th}$ MCMC sample from the posterior distribution. \textcolor{red}{Matt, I think there's an absolute value problem here. Dimorphism can increase of M-F gets more positive or more negative. I've indicated that in the text below, but I could use some help refininig the explanation -YS}
For each trait, the proportion of values in $\boldsymbol{\tau}_k$ that are greater than 0 ($\frac{1}{R} \sum_{r=1}^R I(\tau_{k,r} > 0)$) may indicate that males have gotten relatively larger than females, indicating more pronounced dimorphism. For each trait, the proportion of values in $\boldsymbol{\tau}_k$ that are less than 0 ($\frac{1}{R} \sum_{r=1}^R I(\tau_{k,r} < 0)$) may indicate that females have gotten relatively larger than males, indicating more pronounced dimormphism.

# Results {#sec:results}
The extant data used here consist of a total of `r sum(!is.na(stick$gender))` specimens all with known sex (Table S1).  Of these, there are `r sum(stick$gender == "Female",na.rm = TRUE)` and `r sum(stick$gender == "Male",na.rm = TRUE)` female and male specimens, respectively.

## Missing Data Imputation
\textcolor{red}{WE NEED RESULTS SECTION DESCRIBING THE RESULTS OF AKHIL'S MULTIPLE IMPUTATION, BOTH THE TESTS OF HOW WELL THE MODEL WORKED CLASSIFYING MODEL SAMPLES, AND SUMMARY STATISTICS OF FOSSIL IMPUTATION RESULTS -YS}

@@

## Changes over time
We can also assess the strength of the dimorphism by calculating the proportion of posterior samples that are greater than a prespecified threshold \textcolor{red}{(0.5??, idk what this could be.) -MS}

Table \ref{tab:auc-table} shows the probability that the posterior distribution of the differences in the mean is greater than the posterior distribution of the differences at the first time point.  Values near 0.5 indicate very little difference between the distributions whereas values closer to 0 and 1 indicate changes in the distributions with values of 1 meaning the difference in means has shifted towards males having a larger mean and values of 0 indicating the difference in means has shifted towards larger means in females.  These results are shown graphically in figure \ref{fig:AUC-plot}.

# Acknowledgements {#sec:acknowl}

# Data and Code Statement {#sec:datastatement}
All code for reproducing the analyses in this paper is publicly available at https://github.com/Akhil-Ghosh/SticklebackProject
\textcolor{red}{Is this true? Matt's code is at this GitHub page? -YS}

# References {#sec:references}

# Supplementary Material {#sec:supplementary}
<!-- SCRAP CODE and TEXT. YS moved it down here on 21 October 2024 because it was making it hard for me to see the manuscript.  -->
<!-- THIS FIGURE DOESN'T SHOW MUCH. LET'S CUT FOR SPACE.  -->
```{r fig.cap = "Fossil sample size for each time period",include=FALSE}
ggplot(aes(x = time, y = count), data = fortab) + geom_col() + xlab("Sample")  + scale_x_continuous(breaks = 1:18, labels = 1:18)
```
<!-- I removed the table and replaced it with a figure -->
<!-- % latex table generated in R 4.2.2 by xtable 1.8-4 package -->
<!-- % Mon May  1 11:20:28 2023 -->
<!-- \begin{table}[ht] -->
<!-- \centering -->
<!-- \caption{The number of stickleback fossils at each time point.  Sample size at each time point ranges from a low of 22 to a high of 67.} -->
<!-- \begin{tabular}{rl} -->
<!--   \hline -->
<!-- time & count \\  -->
<!--   \hline -->
<!--   1 & 43 \\  -->
<!--     2 & 41 \\  -->
<!--     3 & 51 \\  -->
<!--     4 & 41 \\  -->
<!--     5 & 46 \\  -->
<!--     6 & 48 \\  -->
<!--     7 & 67 \\  -->
<!--     8 & 55 \\  -->
<!--     9 & 42 \\  -->
<!--    10 & 33 \\  -->
<!--    11 & 37 \\  -->
<!--    12 & 22 \\  -->
<!--    13 & 41 \\  -->
<!--    14 & 43 \\  -->
<!--    15 & 46 \\  -->
<!--    16 & 47 \\  -->
<!--    17 & 56 \\  -->
<!--    18 & 55 \\  -->
<!--    \hline -->
<!-- \end{tabular} -->
<!-- \label{tabn} -->

<!-- Traits such as standard length (stl), anal fin ray (maf), dorsal fin ray (mdf), premaxilla (pmx), and dorsal spine 1 (ds1) demonstrate constant sexual dimorphism with all but the last having learger mean values in male sepcimen. Other traits such as abdominal vertebra (mav), caudal vertebra (mcv), and pelvic spine length (lps) don't appear to have any discernable sexual dimorphism in the means.  Dorsal spine 3 (ds3) exhibits sexual dimorphism that changes over time.  The mean number of dorsal spine 3 begins larger for females for the earliest observed time periods.  Over time, however, this switches and the male specimens exhibit larger mean values for the latest observed time periods.  -->

<!-- \textcolor{red} { -->
<!-- Yoel's interpretation of the three paragraphs above. Male and female G. doryssus differ for some traits, indicating detectable sexual dimorphism. This is evidenced by traits with substantial differences in male and female posterior means (i.e., deviation from 0 in Figure \ref{fig:mu-diff-plot}), supported by posterior probabilities that deviate from 0.5 (Figure \ref{fig:prob0-plot}). For example, males are larger than females by 3-8mm on average through time (stl panel in Figure \ref{fig:mu-diff-plot}), with a posterior probability for that mean difference above 70 percent through time (Figure \ref{fig:prob0-plot}). In contrast, the number of abdominal vertebrae (mav) fluctuates about 0 (Figure \ref{fig:mu-diff-plot}), with posterior probabilities fluctuating about 50 percent through time (Figure \ref{fig:prob0-plot}). Table \ref{tab:prob0-table} in the Supplementary Material summarizes the posterior probabilities that male means differ from female means by trait and sample. -->
<!-- } -->
<!-- The amount of sexual dimporphism over time does not change substantially for a majority of the trait observed here (i.e. cle, ds1, ds2, ect, lps, lpt, mav, mcv, pmx, stl).  Traits that trend away from the mean difference at the first observed time point include dorsal spine 3 (ds3), anal fin ray (maf), and dorsal fin ray (mdf).   -->

<!-- #ds3 -->
<!-- apply(mu_m,2,mean)[1] #2.291846 -->
<!-- apply(mu_f,2,mean)[1] #2.448115  -->
<!-- apply(mu_m,2,mean)[18] #1.264728  -->
<!--   apply(mu_f,2,mean)[18] #1.184534  -->
<!-- apply(mu_m,2,mean)[18] - apply(mu_m,2,mean)[1] #-1.027118  -->
<!-- apply(mu_f,2,mean)[18] - apply(mu_f,2,mean)[1] #-1.263581  -->
<!-- The posterior mean of the mean number of dorsal spine rays at the first observed time point is 2.29 for males and for females it is 2.45.  At the last observed time point, the average number for males and female both dropped to 1.265 and 1.185 for males and females, respectively.  This resulted in a mean number of dorsal spine 3 lost by males of 1.027 and for females it was 1.264.  This resulted in a change of the mean difference from time point 1 to time point 18 of 0.236.  Basically, males and females both lowered the mean number of dorsal spine 3, but females lowered their mean more than the males.   -->

@@

<!-- We also observed a change in the differences between males and females for the number of bones int he anal fin ray (maf) and the number of bones in the dorsal fin ray.  The difference in the means of males versus females for anal fin ray bones changed by 0.152 over the course of the observed time periods.  This change was driven mostly by the change in females.  At the first observed time period the had on average 8.291 bones dropping to 8.170 over the 18 times periods observed. Whereas for males the change in means rose only from  8.518 to 8.549, about 0.0316 bones on average.   -->

<!-- A similar change in the difference of the means over time was observed for the number of bones in the dorsal fin (mdf).  However, unlike maf, this change in sexual dimporhism was driven entirely by the males.  The mean change for females over the 18 observed time periods was only 0.0055 with the average number of dorsal fin bones for females starting at 9.287 and the final observed mean of 9.281 resulting in almost no change in the mean.  For males, the average number of dorsal fin bones at the first time point was about 9.568 dropping to 9.734 over the course of the observed time periods for a gain of 0.166 bones on average.   -->

@@

<!-- \textcolor{red} { -->
<!-- Yoel's interpretation of the five paragraphs above.  -->
<!-- Figure \ref{fig:AUC-plot} shows the posterior probability that the mean male-female difference at time t is different from the male-female difference at time zero. In other words, Figure \ref{fig:AUC-plot} plots whether the sexual dimorphism changes in subsequent samples, always relative to the first sample. The amount of sexual dimporphism over time does not change substantially for a majority of the traits (i.e. cle, ds1, ds2, ect, lps, lpt, mav, mcv, pmx, stl; Figure \ref{fig:AUC-plot}). Traits that change in dimorphism are anal fin ray (maf), and dorsal fin ray (mdf), and dorsal spine 3 (ds3).} -->
